---
title: "SDS 315 Homework 7"
author: "Yash Gupta - yg9449"
date: "2025-04-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(scipen = 999)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(mosaic)
library(MatchIt)
```

Link to github respository with Rmd file: <https://github.com/ygnewyork/SDS_HW7>

## Question 1 - Armfolding

```{r}

arms <- read_csv("armfold.csv")

```

### Part A

```{r}

male_count <- sum(arms$Sex == "Male")
female_count <- sum(arms$Sex == "Female")

male_left <- sum(arms$Sex == "Male" & arms$LonR_fold == 1)
female_left <- sum(arms$Sex == "Female" & arms$LonR_fold == 1)

male_prop <- male_left / male_count
female_prop <- female_left / female_count

```

The number of males in the dataset is 106, and number of females is 111. The sample proportion of males who folded their left arm on top is about 0.47, and for females is about 0.42.

### Part B

```{r}
diff_prop <- male_prop - female_prop
```

The observed difference in proportions between the two groups who folded their left arm on top (males - females) is about 0.05.

### Part C

```{r}

lefts <- c(male_left, female_left)
counts <- c(male_count, female_count)

result <- prop.test(lefts, counts)  
confint_result <- result$conf.int

se <- sqrt((male_prop * (1 - male_prop) / male_count) + (female_prop * (1 - female_prop) / female_count))

z_score <- 1.96

moe <- se * z_score

lower_bound <- (diff_prop - moe)
upper_bound <- (diff_prop + moe)

```

The result for the 95% confidence interval using conf.int on prop.test for the difference in proportions between males and females is (-0.09, 0.19). The hand-calculated method was found by calculating the standard error by doing the formula sqrt(male_prop \* (1 - male_prop) / male_count + female_prop \* (1 - female_prop) / female_count) and using a z_score of 1.96 as we're calculating a 95% confidence interval, then finding the MOE by doing se \* z and then finding the lower and upper bounds by subtracting and adding the MOE to the observed difference of proportions, resulting in interval bounds (-0.08, 0.18).

### Part D

If we were to repeat our sampling process and calculate the confidence interval many, many times, then we would expect that about 95% of those intervals would contain the true population parameter.

### Part E

Standard Error above represents the amount of variability we could expect to see from the difference between male and female sample proportions for if the left arm would be on top if we were to draw new samples of the same size of the same population many many times. We are measuring the variability (standard deviation) of how the sample proportion varies from the true proportion.

### Part F

Sampling distribution represents the probability distribution of the difference between the sample proportions for males and females across many repetitions of this study. The items varying from sample to sample would be the individuals chosen, the sample proportions, and the difference between these proportions. The things that would stay the same would be the underlying populations, the true population parameters, and the sample sizes.

### Part G

The Central Limit Theorem is used to justify the use of the normal distribution to approximate the sampling distribution of the difference in sample proportions. The CLT says that if you take large enough samples, the sampling distribution of the sample proportion would be an approximately normal (bell-shaped) distribution, even if the original population isn't itself normally distributed.

### Part H

We can't reject the claim, as the interval (-0.01, 0.30) contains the value of 0, indicating that the study results do indicate that the true proportion could indicate no sex difference in arm folding.

### Part I

If repeated many times, the confidence interval would be different as we would have sampling variability, leading to different samples of potentially more males/females who fold their arms another way, or fewer. Since these samples are different, the sample proportions would also vary, and the confidence intervals that use these proportions would also vary as a result. For all of these intervals, however, if you were to repeat them many, many times, and collect all the different 95% confidence intervals from each sample, 95% of the intervals would contain the true difference in arm-folding proportions between males and females in the entire university student population.

## Question 2 - Turnout

```{r}

turnout <- read_csv("turnout.csv")

```

### Part A

```{r}

call_count <- sum(turnout$GOTV_call == 1) 
nocall_count <- sum(turnout$GOTV_call == 0)
  
voted_1998_call <- sum(turnout$voted1998 == 1 & turnout$GOTV_call == 1)
voted_1998_nocall <- sum(turnout$voted1998 == 1 & turnout$GOTV_call == 0)

prop_1998_call <- voted_1998_call / call_count
prop_1998_nocall <- voted_1998_nocall / nocall_count

votes <- c(voted_1998_call, voted_1998_nocall)
totals <- c(call_count, nocall_count)

diff <- prop.test(votes, totals)  
diff
```

The proportion of those receiving a GOTV call who voted in 1998 is 0.648, and the sample proportion of those not receiving a GOTV call who voted in 1998 is 0.444. The large sample 95% confidence interval for the difference in these two proportions is (0.141, 0.266).

### Part B

```{r}

tab_gotv_96 <- table(turnout$GOTV_call, turnout$voted1996)
prop_test_gotv96 <- prop.test(tab_gotv_96)

print("95% Confidence Interval for Voting in 1996 and GOTV calls")
prop_test_gotv96

tab_v96_v98 <- table(turnout$voted1998, turnout$voted1996)
prop_test_v96_v98 <- prop.test(tab_v96_v98)

print("95% Confidence Interval for voting in 1996 and voting in 1998")
prop_test_v96_v98

```

For the voted1996 variable, it's a confounder on the causal effect of the GOTV call affecting the likelihood that a person voted in 1998, as it increases the likelihood of both of those variables. We can see this through calculating the prop.test 95% confidence intervals and seeing that since they both don't contain zero, that voting in 1996 increases your chance of receiving a call AND voting in 1998. Since this variable impacts both of these, it's a confounder that needs to be accounted for when looking for the causal relationship of just the call on 1998 voting likelihood.

```{r}

tab_gotv_majorpty <- table(turnout$GOTV_call, turnout$MAJORPTY)
prop_test_gotv_majorpty <- prop.test(tab_gotv_majorpty)

print("95% Confidence Interval for being registered to a major party and GOTV calls")
prop_test_gotv_majorpty

tab_v98_majorpty <- table(turnout$voted1998, turnout$MAJORPTY)
prop_test_v98_majorpty <- prop.test(tab_v98_majorpty)

print("95% Confidence Interval for being registered to a major party and voting in 1998")
prop_test_v98_majorpty

```

For the MAJORPTY variable, it's a confounder on the causal effect of the GOTV call affecting the likelihood that a person voted in 1998, as it increases the likelihood of both of those variables. We can see this through calculating the prop.test 95% confidence intervals and seeing that since they both don't contain zero, that being registered to a major party increases your chance of receiving a call AND voting in 1998. Since this variable impacts both of these, it's a confounder that needs to be accounted for when looking for the causal relationship of just the call on 1998 voting likelihood.

```{r}
boxplot(AGE ~ GOTV_call, data = turnout,
        names = c("No GOTV Call", "GOTV Call"),
        main = "Age Distribution by Get Out The Vote Calls",
        xlab = "GOTV_call",
        ylab = "AGE")

print("95% Confidence Interval between mean of ages and GOTV calls")
age_call <- t.test(AGE ~ GOTV_call, data = turnout)
age_call

boxplot(AGE ~ voted1998, data = turnout,
        names = c("Did Not Vote 1998", "Voted 1998"),
        main = "Age Distribution by Voting in 1998",
        xlab = "Voted 1998",
        ylab = "AGE")

print("95% Confidence Interval between mean of ages and voting in 1998")
age_vote <- t.test(AGE ~ voted1998, data = turnout)
age_vote

```

For the AGE variable, it's a confounder on the causal effect of the GOTV call affecting the likelihood that a person voted in 1998, as it increases the likelihood of both of those variables. We can see this through identifying the difference on the boxplots, where the age median is significantly higher (also established by the difference in the t.test 95% confidence intervals for the age mean not containing the value of zero) for those who received a call and for those who voted in 1998. Since this variable impacts both of these, it's a confounder that needs to be accounted for when looking for the causal relationship of just the call on 1998 voting likelihood.

### Part C

```{r}

vote_match = matchit(GOTV_call ~ AGE + MAJORPTY + voted1996, data = turnout, ratio=5)

votes_matched <- match.data(vote_match)

print("Matched confidence intervals for GOTV calls and voted in 1996")
match_call_96 <- table(votes_matched$GOTV_call, votes_matched$voted1996)
prop.test(match_call_96)

print("Matched confidence intervals for GOTV calls and is registered to a major party")
match_call_pty <- table(votes_matched$GOTV_call, votes_matched$MAJORPTY)
prop.test(match_call_pty)

print("Matched confidence intervals for GOTV calls and mean ages")
t.test(AGE ~ GOTV_call, data = votes_matched)

```

As shown from the 95% confidence intervals, all of them contain the value zero for the 3 confounding variables of voting in 1996, being registered to a major party, and ages, which shows that the dataset is matched evenly between the groups of receiving or not receiving a call with a ratio of 5 control cases for each treated case.

```{r}
call_count_matched <- sum(votes_matched$GOTV_call == 1) 
nocall_count_matched <- sum(votes_matched$GOTV_call == 0)
  
voted_1998_call_matched <- sum(votes_matched$voted1998 == 1 & votes_matched$GOTV_call == 1)
voted_1998_nocall_matched <- sum(votes_matched$voted1998 == 1 & votes_matched$GOTV_call == 0)

prop_1998_call_matched <- voted_1998_call_matched / call_count_matched
prop_1998_nocall_matched <- voted_1998_nocall_matched / nocall_count_matched

votes_count_matched <- c(voted_1998_call_matched, voted_1998_nocall_matched)
totals_matched <- c(call_count_matched, nocall_count_matched)

diff_matched <- prop.test(votes_count_matched, totals_matched)  
diff_matched
```

The proportion of those receiving a GOTV call who voted in 1998 with the matched data is 0.648, and the sample proportion of those not receiving a GOTV call who voted in 1998 with the matched data is 0.569. The large sample 95% confidence interval for the difference in these two proportions is (0.010, 0.147). Since this interval does not contain the value zero, it means that the GOTV call does statistically significantly (at a 0.95 level) increase the likelihood of voting in the 1998 election, by a couple of percentage points.
